The Community Land Model, version 5.1 \\ One-at-a-time Parameter Perturbation Ensemble

### 2023-05-11 ms/ms.tex
edits
from overleaf

\title{The Community Land Model, version 5: \\ One-at-a-time Parameter Perturbation Ensemble}

\authors{D. Kennedy \affil{1}, Katherine Dagon\affil{1}, David M. Lawrence\affil{1}}
%R.A. Fisher
%B.M. Sanderson
%C.D. Koven
%A.L.S. Swann
%C.M. Zarakas
%L.R. Hawkins
%F. Hoffman
%N. Collier
%Y. Luo
%X. Lu
%S. Levis
%S.C. Swenson
%K.W. Oleson
%E. Kluzek
%J. Liu

@@

Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within the land component of Earth System Models.
Understanding how these variables respond to carbon dioxide concentration, and how we expect carbon dioxide concentration to evolve over time, is at the center of climate change research.  
Our certainty in climate model projections varies by domain and generally decreases with extended time horizons.
Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, we would expect predictions to become more skillful over time.

@@

One such fundamental model development step is parameter sensitivity testing and/or parameter optimization.
Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017}.
When applied systematically, the result is a parameter perturbation ensemble (PPE), alternatively termed a perturbed physics ensemble.
PPEs have played an important role in assessing projection uncertainty in climate models.
They also form a basis for automated model calibration.
Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes. 
In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, how it was produced, and its potential applications.

@@

We used k-means clustering to identify groups of grid cells with similar climates based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase. We selected one representative grid cell from each cluster to stand in for the entire cluster. The representative grid cell is whichever is located nearest the cluster centroid in climate space. The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a `coarse' grid. A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.

@@

\subsection{Experimental Design}
In the preparation for this experiment, 188 CLM-BGC parameters were identified. We decided to vary each parameter independently, to a low and high value. To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter. In some cases literature values were directly utilized, but in the majority of cases, expert judgment was used. The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz.

@@

\subsection{Parameters} 
\begin{landscape}

 \begin{table}[h]
 \caption{Some key parameters \\
 Right now I'm sorting them by domain, should I sort them alphabetically by parameter name instead?}
 \centering
 \begin{tabular}{l l c}
 \hline
  Parameter  & Description & Model Domain \\
 \hline
d\_max & Dry surface layer parameter & Sensible, latent heat and momentum fluxes \\
frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
fff & Decay factor for fractional saturated area & Hydrology \\
liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology \\
maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology \\
medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis \\
jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity (LUNA) \\
jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity (LUNA) \\
slatop & Specific leaf area at top of canopy & Photosynthetic capacity (LUNA) \\
wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity (LUNA) \\
kmax & Plant segment max conductance & Plant hydraulics \\
krmax & Root segment max conductance & Plant hydraulics \\
psi50 & Water potential at 50\% loss of conductance & Plant hydraulics \\
nstem & Stem number & Biomass heat storage \\
lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration \\
froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation \\
leafcn & Leaf C:N & Carbon and nitrogen allocation \\
leaf\_long & Leaf longevity & Vegetation phenology and turnover \\
cpha & Activation energy for cp & Acclimation parameters \\
jmaxhd & Deactivation energy for jmax & Acclimation parameters \\
kcha & Activation energy for kc & Acclimation parameters \\
lmrha & Activation energy for lmr & Acclimation parameters \\
lmrhd & Deactivation energy for lmr & Acclimation parameters \\
tpuha & Activation energy for tpu & Acclimation parameters \\
tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters \\
vcmaxha & Activation energy for vcmax & Acclimation parameters \\
vcmaxhd & Deactivation energy for vcmax & Acclimation parameters \\
 \hline
 \end{tabular}
 \end{table}
\end{landscape}

@@

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
stronger the relationship between output from the full grid and the sparsegrid. Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sggpp}). We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
\begin{figure}[h]
\centering
\includegraphics[width=35pc]{../figs/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sg}
\end{figure}

Most of the parameter perturbations had a relatively small impact on any given output variable. For example, the distributions of GPP and NBP are concentrated around the default parameterization with long tails, indicating that a small proportion of parameters have large effects (Figure \ref{fig:violins}). One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death. This simulation is likely unrealistic, indicating that the perturbation range was too large, but possibly also that the model struggles to acclimate to warmer soils. This represents one of the valuable outcomes of an extensive PPE, exposing unexpected model behavior, which may belie brittle parameterizations or bugs.

>>>>

\title{The Community Land Model, version 5.1 \\ One-at-a-time Parameter Perturbation Ensemble}

\authors{D. Kennedy\affil{1} and many more}

@@

Water availability, land temperature extremes, fire risk, and crop productivity will all see impacts from climate change, and are among the many variables represented within the land component of Earth System Models.
Understanding how these variables respond to CO$_2$ concentration, and how we expect CO$_2$ concentration to evolve over time, is at the center of climate change research.  
Our certainty in climate model projections varies by domain and generally decreases with extended time horizons.
Centennial-scale estimates of the cumulative terrestrial carbon sink have been especially challenging, with high uncertainty persisting across model generations \cite{friedlingstein2014,arora2020}. 
A portion of this uncertainty is irreducible, inherent to the challenge of predicting vegetation dynamics in a novel climate \cite{bonan2018}.
But with the ever-increasing observational basis from remote sensing, meteorological stations, flux towers, and field campaigns, we might expect predictions to become more skillful across subsequent model generations.

@@

One such fundamental model development step is parameter sensitivity testing and/or parameter optimization.
Because parameter values are uncertain, evaluating the effect of parameter values on model output is critical to the model development process \cite{hourdin2017}.
When applied systematically, the result is a parameter perturbation ensemble (PPE), alternatively termed a perturbed physics ensemble.
PPEs have played an important role in assessing projection uncertainty in climate models (Hawkins 2019 JAMES).
They also form a basis for automated model calibration.
Our goal in this project is to systematize parameter perturbation activity within our modeling framework, and develop the necessary tools and datasets to efficiently test parameter effects across the full suite of land model processes. 
In doing so, we have generated a large PPE, comprising thousands of parameter sensitivity tests.
In this paper we present that dataset, how it was produced, and its potential applications.

@@

We used k-means clustering to identify groups of grid cells with similar dynamics based on a 2$^\circ$ transient simulation (1850-2014) using the CLM-PPE codebase. We selected one representative grid cell from each cluster to stand in for the entire cluster. The representative grid cell is whichever is located nearest the cluster centroid in climate space. The set of representative grid cells comprise a `sparsegrid', which are used in lieu of a `coarse' grid. A paint-by-number approach is used to recompose mapped output and compute global means, where the output from the representative grid cell is substituted for all members of the cluster cohort.

@@

\subsection{Experimental Design}
In the preparation for this experiment, 188 CLM-BGC parameters were identified. We decided to vary each parameter independently, to a low and high value. To define parameter ranges we created an online spreadsheet and solicited domain-area experts to provide a minimum and maximum value for each parameter. In some cases literature values were directly utilized, but in the majority of cases, expert judgment was used. The spreadsheet, with literature references and parameter descriptions is available online and in appendix zqz. In some cases parameters could not be perturbed independently, because they are required to sum to 1. In the case of nitrogen fixation costs, we opted to perturbed the parameters independently, but also as a group (`KCN'), to enforce nitrogen limitation in lieu of switching from one uptake pathway to another.

@@

\subsection{Parameters} 
\begin{landscape}
 \begin{table}[h]
 \caption{Some key parameters}
 \centering
 \begin{tabular}{l l c}
 \hline
  Parameter  & Description & Model Domain \\
 \hline
d\_max & Dry surface layer parameter & Sensible, latent heat and momentum fluxes \\
frac\_sat\_soil\_dsl\_init & Fraction of saturated soil at which DSL initiates & Sensible, latent heat and momentum fluxes \\
fff & Decay factor for fractional saturated area & Hydrology \\
liq\_canopy\_storage\_scalar & Canopy-storage-of-liquid-water parameter & Hydrology \\
maximum\_leaf\_wetted\_fraction & Maximum leaf wetted fraction & Hydrology \\
medlynintercept & Medlyn intercept of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
medlynslope & Medlyn slope of conductance-photosynthesis relationship & Stomatal resistance and photosynthesis \\
tpu25ratio & Ratio of tpu25top to vcmax25top & Stomatal resistance and photosynthesis \\
jmaxb0 & Baseline proportion of nitrogen allocated for electron transport & Photosynthetic capacity (LUNA) \\
jmaxb1 & Response of electron transport rate to light & Photosynthetic capacity (LUNA) \\
slatop & Specific leaf area at top of canopy & Photosynthetic capacity (LUNA) \\
wc2wjb0 & Baseline ratio of wc:wj & Photosynthetic capacity (LUNA) \\
kmax & Plant segment max conductance & Plant hydraulics \\
krmax & Root segment max conductance & Plant hydraulics \\
psi50 & Water potential at 50\% loss of conductance & Plant hydraulics \\
nstem & Stem number & Biomass heat storage \\
lmr\_intercept\_atkin & Intercept in the calculation of leaf maintenance respiration& Plant respiration \\
froot\_leaf & Allocation parameter: new fine root C per new leaf C & Carbon and nitrogen allocation \\
leafcn & Leaf C:N & Carbon and nitrogen allocation \\
leaf\_long & Leaf longevity & Vegetation phenology and turnover \\
cpha & Activation energy for cp & Acclimation parameters \\
jmaxhd & Deactivation energy for jmax & Acclimation parameters \\
kcha & Activation energy for kc & Acclimation parameters \\
lmrha & Activation energy for lmr & Acclimation parameters \\
lmrhd & Deactivation energy for lmr & Acclimation parameters \\
tpuha & Activation energy for tpu & Acclimation parameters \\
tpuse\_sf & Scale factor for tpu entropy term & Acclimation parameters \\
vcmaxha & Activation energy for vcmax & Acclimation parameters \\
vcmaxhd & Deactivation energy for vcmax & Acclimation parameters \\
 \hline
 \end{tabular}
 \end{table}
\end{landscape}

@@

Choosing the number of clusters to generate the sparsegrid involved balancing the computational savings against representational fidelity. Generally the more clusters, the 
stronger the relationship between output from the full grid and the sparsegrid. Accuracy of global photosynthesis could be achieved with a relatively small number of clusters, with R$_2>$0.95 achieved with only 200 clusters, and RMSE$<$1PgC requiring approximately 700 clusters (Figure \ref{fig:sg}). We chose 400 clusters, based on ILAMB scores of a wide variety of output variables (Supp Figure \ref{supp:ilamb})
\begin{figure}[h]
\centering
\includegraphics[width=35pc]{../figs/sparsegrid_gpp.pdf}
\caption{Sparsegrid vs fullgrid (2$^{\circ}$ resolution) global annual GPP across the last thirty years of a transient CLM5.1 simulation. We opted for 400 clusters to balance computational cost against representativeness.}
\label{fig:sg}
\end{figure}

Most of the parameter perturbations had a relatively small impact on any given output variable. For example, the distributions of GPP and NEP are concentrated around the default simulation with long tails, indicating that a small proportion of parameters have large effects (Figure \ref{fig:violins}). One perturbation, reducing the heat capacity of sand by 20\%, proved destructive in the future climate scenario, resulting in inhospitably hot soil conditions and widespread plant death. This simulation is likely unrealistic, indicating that the perturbation range was too large, but possibly also that the model struggles to acclimate to warmer soils. This represents one of the valuable outcomes of an extensive PPE, exposing unexpected model behavior, which may belie brittle parameterizations or bugs.
